GADTs, Functoriality, Parametricity: Pick Two

### 2021-12-03 20-gadts/21-LSFA/GADT-story.tex
final LSFA paper for uploading

\noindent
This sequence of syntactic generalizations suggests that a
corresponding sequence of semantic generalizations might also hold.
However, this is not the case. Specifically, the semantic equivalence
between Church encodings and fixpoint representations that holds for
ADTs and nested types need not hold for GADTs. In a language
supporting GADTs we must therefore choose whether to represent data
types as their Church encodings or as fixpoints.  The main result of
this paper shows that the choice of representation of data types in a
language supporting GADTs determines whether or not that language can
have parametric models~\cite{rey83}. It thus determines whether or not
GADTs can enjoy consequences of parametricity such as representation
independence~\cite{adr09,dnb12}, equivalences between
programs~\cite{hd11}, deep induction principles~\cite{jgj21s,jp20},
and useful (``free'') theorems about programs derived from their types
alone~\cite{wad89}.  This result is unintuitive, novel, and
surprising.

>>>>

\noindent
This sequence of syntactic generalizations suggests that a
corresponding sequence of semantic generalizations might also hold.
However, this is not the case. Specifically, the semantic equivalence
between Church encodings and fixpoint representations that holds for
ADTs and nested types need not hold for GADTs. In a language
supporting GADTs we must therefore choose whether to represent data
types as their Church encodings or as fixpoints.  The main result of
this paper shows that the choice of representation of data types in a
language supporting GADTs determines whether or not that language can
have parametric models~\cite{rey83}. It thus determines whether or not
GADTs can enjoy consequences of parametricity such as representation
independence~\cite{adr09,dnb12}, equivalences between
programs~\cite{hd11}, deep induction principles~\cite{jg22,jp20},
and useful (``free'') theorems about programs derived from their types
alone~\cite{wad89}.  This result is unintuitive, novel, and
surprising.
